Some lessons after the formalization of the ctm approach to forcing \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}

### 2022-05-06 JSL/forcing.tex
New discipline
More on relativization. Extensions to Isabelle/ZF. Few words on forcing

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
\section{Forcing}
\label{sec:forcing}

Prerequisite: a set $\formula$ of codes for first order formulas, and
the (recursive) definition of the Tarski satisfaction relation in $\ZF$.

Let $\lb \PP, {\preceq} ,\1\rb \in M$ be a forcing notion. Given $G\sbq \PP$, we have
$M[G]\defi \{ \val(\PP,G,\punto{a}) : \punto{a}\in M \}$.

The following form of the Forcing Theorems  is the one
that we formalized.
\begin{theorem}
  There exists a function  $\forceisa: \formula \to  \formula$
  such that for every
  $\phi$ and $\punto{a}_0,\dots,\punto{a}_n\in M$,
  \begin{enumerate}
  \item (Truth Lemma) for every $M$-generic $G$,
    \[
      M[G], [\val(\PP,G,\punto{a}_0),\dots,\val(\PP,G,\punto{a}_n)]
      \models \phi
    \]
    is equivalent to 
    \[
      \exists p\in G.\ \; M, [p,\PP,\preceq,\1, \punto{a}_0,\dots,\punto{a}_n]  \models
      \forceisa(\phi).\]

@@

Note: $\forceisa$ was formalized as a class-function (i.e., of type
$\tyi\Rightarrow\tyi$), to take advantage of the Isabelle's
conveniences for primitive recursion.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
\section{Forcing}
\label{sec:forcing}
Let $\lb \PP, {\preceq} ,\1\rb \in M$ be a forcing notion. Given $G\sbq \PP$, we have
$M[G]\defi \{ \val(\PP,G,\punto{a}) : \punto{a}\in M \}$.

The following form of the Forcing Theorems  is the one
that we formalized.
\begin{theorem}
  There exists a function  $\forceisa:: \tyi \fun  \tyi$
  such that for every
  $\phi\in\formula$ and $\punto{a}_0,\dots,\punto{a}_n\in M$,
  \begin{enumerate}
  \item (Definability) $\forceisa(\phi)\in\formula$;
  \item (Truth Lemma) for every $M$-generic $G$,
    \[
      M[G], [\val(\PP,G,\punto{a}_0),\dots,\val(\PP,G,\punto{a}_n)]
      \models \phi
    \]
    is equivalent to 
    \[
      \exists p\in G.\ \; M, [p,\PP,\preceq,\1, \punto{a}_0,\dots,\punto{a}_n]  \models
      \forceisa(\phi).\]

@@

We followed the new Kunen's book to define
$\forceisa$.  Forcing for atomic formulas is described as a mutual
recursion
%% \begin{multline*}
%%   \forceseq (p,t_1,t_2) \defi 
%%   \forall s\in\dom(t_1)\cup\dom(t_2).\ \forall q\pleq p .\\
%%   \forcesmem(q,s,t_1)\lsii \forcesmem(q,s,t_2)
%% \end{multline*}
%% \begin{multline*}
%%   \forcesmem(p,t_1,t_2) \defi  \forall v\pleq p. \ \exists q\pleq v.\\  
%%   \exists s.\ \exists r\in \PP .\ \lb s,r\rb \in  t_2 \land q
%%   \pleq r \land \forceseq(q,t_1,s)
%% \end{multline*}
but then \cite[p.~257]{kunen2011set} it is cast as a single
recursively defined function $\frcat$ over the wellfounded relation
$\isatt{frecR}$ on tuples $\lb \mathit{ft},t_1,t_2,p\rb$ (where
$\mathit{ft}\in\{0,1\}$ indicates the type of the atomic formula being
forced). Forcing for general formulas is defined by recursion on the
datatype $\formula$. Details on the implementation and proofs of the
Forcing Theorems have been spelled out in our
\cite{2020arXiv200109715G}.

It is to be noted that application of the Forcing theorems do not
require any extra Replacement instances on $M$.

### 2022-05-06 JSL/non-absolute.tex
More on relativization. Extensions to Isabelle/ZF. Few words on forcing

The treatment of relativization/internalization described in the
previous section was enough for the Paulson's treatment of
constructibility. This is the case because essentially all the
concepts in the way of proving the consistency of $\AC$ are
absolute, and treatment of relational versions and relativized notions
could be minimized after proving the relevant absoluteness results:
For example, the lemma \isatt{Union{\uscore}abs},
\[
  M(A) \implies M(z) \implies \isatt{big{\uscore}union}(M, A, z) \longleftrightarrow z = \union
  A
\]
proved under the assumption that $M$ is transitive nonempty. Working in this
way with powersets, cardinalities, and the like would be
unfeasible.

In order to cope with this, we added the missing step from the
literature consisting of relative versions of the various non-absolute
functions and limited automatic facilities that define, state, and
prove the needed concepts. For instance, the 
$\isatt{cardinal}::\tyi \fun \tyi$ function is defined in
\session{Isabelle/ZF}, and the commands
\begin{isabelle}
  \isacommand{relativize}\isamarkupfalse%
  \ \isakeyword{functional}\ {\isachardoublequoteopen}cardinal{\isachardoublequoteclose}\ {\isachardoublequoteopen}cardinal{\isacharunderscore}{\kern0pt}rel{\isachardoublequoteclose}\ \isakeyword{external}\isanewline
  \isacommand{relationalize}\isamarkupfalse%
  \ {\isachardoublequoteopen}cardinal{\isacharunderscore}{\kern0pt}rel{\isachardoublequoteclose}\ {\isachardoublequoteopen}is{\isacharunderscore}{\kern0pt}cardinal{\isachardoublequoteclose}\isanewline
  \isacommand{synthesize}\isamarkupfalse%
  \ {\isachardoublequoteopen}is{\isacharunderscore}{\kern0pt}cardinal{\isachardoublequoteclose}\ \isakeyword{from{\isacharunderscore}{\kern0pt}definition}\ \isakeyword{assuming}\ {\isachardoublequoteopen}nonempty{\isachardoublequoteclose}%
\end{isabelle}
define the relative cardinal function
$\isatt{cardinal{\uscore}rel}::(\tyi \fun \tyo) \fun \tyi \fun\tyi$,
its relational version $\isatt{is{\uscore}cardinal}$ that encodes the
statement $|x|^M = z$, the
internalized formula \isatt{is{\uscore}cardinal{\uscore}fm} whose
satisfaction by a set is equivalent to the relational version, and the
proof of the previous statements.

>>>>

The treatment of relativization/internalization described in the
previous section was enough for the Paulson's treatment of
constructibility. This is the case because essentially all the
concepts in the way of proving the consistency of $\AC$ are
absolute, and treatment of relational versions and relativized notions
could be minimized after proving the relevant absoluteness results:
For example, the lemma \isatt{Union{\uscore}abs},
\[
  M(A) \implies M(z) \implies \isatt{big{\uscore}union}(M, A, z) \longleftrightarrow z = \union
  A
\]
proved under the assumption that $M$ is transitive nonempty. Working
in this relational 
way with powersets, cardinalities, and the like would be
unfeasible. As such, cardinal arithmetic was not put in relative form
in \session{ZF-Constructible}.

\subsection{Tools for relativization}
\label{sec:tools-relativization}
In order to cope with this, we added the missing step from the
literature consisting of relative versions of the various non-absolute
functions and programmed in ML limited automatic facilities that
define the needed concepts, and  state and
prove the requisite lemmas. For instance, the 
$\isatt{cardinal}::\tyi \fun \tyi$ function is defined in
\session{Isabelle/ZF}, and the commands
\begin{isabelle}
  \isacommand{relativize}\isamarkupfalse%
  \ \isakeyword{functional}\ {\isachardoublequoteopen}cardinal{\isachardoublequoteclose}\ {\isachardoublequoteopen}cardinal{\isacharunderscore}{\kern0pt}rel{\isachardoublequoteclose}\ \isakeyword{external}\isanewline
  \isacommand{relationalize}\isamarkupfalse%
  \ {\isachardoublequoteopen}cardinal{\isacharunderscore}{\kern0pt}rel{\isachardoublequoteclose}\ {\isachardoublequoteopen}is{\isacharunderscore}{\kern0pt}cardinal{\isachardoublequoteclose}\isanewline
  \isacommand{synthesize}\isamarkupfalse%
  \ {\isachardoublequoteopen}is{\isacharunderscore}{\kern0pt}cardinal{\isachardoublequoteclose}\ \isakeyword{from{\isacharunderscore}{\kern0pt}definition}\ \isakeyword{assuming}\ {\isachardoublequoteopen}nonempty{\isachardoublequoteclose}%
\end{isabelle}
define the relative cardinal function
$\isatt{cardinal{\uscore}rel}::(\tyi \fun \tyo) \fun \tyi \fun\tyi$
(denoted  $|\cdot|^M$, as expected),
its relational version $\isatt{is{\uscore}cardinal}$, the
internalized formula \isatt{is{\uscore}cardinal{\uscore}fm} whose
satisfaction by a set is equivalent to the relational version, and
prove the previous statement (analogous to (\ref{eq:sats_big_union_fm})).
The proof that $\isatt{is{\uscore}cardinal}(M,x,z)$  encodes the
statement $|x|^M = z$ must still be done by hand, since the definition
of $\isatt{cardinal{\uscore}rel}$ already involves some tacit
absoluteness result (“\textit{the least $\ka \in \Ord$ such that \dots}” instead
of “\textit{the least $\ka \in \Ord^M$ such that \dots}”, and the like).

\subsection{Extension of Isabelle/ZF}
\label{sec:extension-isabellezf}
We extended \cite{Delta_System_Lemma-AFP} the material formalized in
Isabelle, from basic results involving function spaces and the
definition of cardinal exponentiation, to a treatment of cofinality
and the Delta System Lemma for $\omega_1$-families. We also included a
concise treatment of the axiom of Dependent Choices $\DC$ and the
general version of Rasiowa-Sikorski Lemma \cite{2018arXiv180705174G}
and a choiceless one for countable preorders.

This material was subsequently put in relative form in our formal
development on transitive class models \cite{Transitive_Models-AFP}
using as an aid the tools from
Section~\ref{sec:tools-relativization}. We also relativized many
original theories appearing in \session{ZF}, including the
fundamentals of cardinal arithmetic, the cumulative hierarchy, and the
definition of the $\ale{}$ function.

### 2022-05-06 JSL/proof_assists_isabelle.tex
More on relativization. Extensions to Isabelle/ZF. Few words on forcing

\subsection{Isabelle metalogic $\Meta$}

@@

The type of propositions $\prop$ equipped with a binary operation
$\implies$ (“meta-implication”) that is used to represent the object
logic rules. As an example, the axiomatization of first-order logic
postulates a type $\tyo$ of booleans, and Modus Ponens
% https://isabelle.in.tum.de/dist/library/FOL/FOL/IFOL.html#IFOL.mp|axiom
is written as
\[
  [P\limp Q] \implies ([P] \implies [Q]).
\]
The square brackets (which are omitted in Isabelle theories) represent
an injection from $\tyo$ into $\prop$. % ($[P]$ can be read as “$P$
% holds”)
A consequence of this representation is that every formula of
the object logic appears as atomic for $\Meta$.

%% Quantification is handled in the meta-level using a functional $\ALL$
%% with polymorphic type $(\alpha \fun \prop) \fun \prop$. 
Types in Isabelle are organized into \emph{classes} and \emph{sorts};
for ease of exposition, we will omit the former.  The axiomatization
of first-order logic postulates a sort $\{\type{term}\}$ (of
“individuals,” or elements of a first-order universe of discourse) and
stipulates that every further type variable $\alpha$ must be of that
sort. In particular, Isabelle/ZF only postulates one new type $\tyi$
(“sets”) of sort $\{\type{term}\}$. Hence, from the type of the universal
quantifier functional $\forall :: (\alpha \fun \tyo) \fun \tyo$ it
follows that it may only be applied to predicates with a variable of
type $\tyi$. This ensures that the object logic is effectively
first-order.

@@

\subsection{Isabelle/ZF}

@@

Isabelle/ZF reaches Hessenberg's $|A|\cdot|A| = |A|$. Our decision
(during 2017) to
use this assistant was triggered by its constructibility
library \citep{paulson_2003},
% https://isabelle.in.tum.de/dist/library/ZF/ZF-Constructible/
which contains the development of $L$, the proof that it satisfies
$\AC$, and a version of the Reflection Principle. The latter was
actually encoded as a series of instructions to Isabelle automatic
proof tools that would prove each particular instance of reflection:
This is an example of what was said at the end of the previous
section.

@@

\subsection{Internalized formulas}

\session{ZF-Constructible} defines the set $\formula$ of codes for
first-order formulas. These, alongside with lists, are instances of
Isabelle/ZF treatment of inductively defined datatypes; induction
and recursion theorems for them are proved automatically.

To avoid problems related to the binding of variables, de Bruijn
indices are used instead. The satisfaction predicate
$\isatt{sats}::\tyi\fun\tyi\fun\tyi\fun\tyo$ takes as arguments a set $M$, a list
$\mathit{ms}\in\isatt{list}(M)$ for the assignment of free indices,
and $\phi\in\formula$, and it is written
$M,\mathit{ms}\models\phi$ in our formalization. This completes the
description of the fourth and last formal layer of the development.

Internalized formulas for most (but not all) of the relational
concepts can be obtained by guiding the automatic tactics. Actually,
great many of the concepts in \session{ZF-Constructible} where
internalized by hand; this is the case for union,
%% \begin{multline*}
%%   \isatt{big{\uscore}union{\uscore}fm}(A,z) \equiv \\
%%   ((\cdot\forall\cdot\cdot0 \in succ(z)\cdot \longleftrightarrow
%%   (\cdot\exists\cdot\cdot0 \in succ(succ(A))\cdot \land \cdot1 \in
%%   0\cdot\cdot\cdot)\cdot\cdot)
%% \end{multline*}
for which we have the following satisfaction lemma:
% https://isabelle.in.tum.de/dist/library/ZF/ZF-Constructible/L_axioms.html#L_axioms.sats_big_union_fm|fact
\begin{multline*}
  x \in \omega \implies y \in \omega \implies \mathit{env} \in \isatt{list}(A)
  \implies \\
  (A, \mathit{env} \models \isatt{big{\uscore}union{\uscore}fm}(x,y)%\cdot\union x \isatt{ is } y\cdot
  ) \longleftrightarrow  \isatt{big{\uscore}union}(\isatt{\#\#} A, \isatt{nth}(x,
  \mathit{env}), \isatt{nth}(y, \mathit{env}))
\end{multline*}
Note that $x$ and $y$ above are de Bruijn indices,
$\isatt{nth}(x,\mathit{env})$ is the $x$th element of $\mathit{env}$
and $\isatt{\#\#}A::\tyi\fun \tyo$ is the class corresponding to the
set $A::\tyi$.

>>>>

\subsection{Isabelle metalogic $\Meta$}
\label{sec:isabelle-metalogic-meta}

@@

The type of propositions $\prop$ equipped with a binary operation
$\implies$ (“meta-implication”) and a universal “meta-quantifier”
$\ALL$, that are used to represent the object
logic rules. As an example, the axiomatization of first-order logic
postulates a type $\tyo$ of booleans, and Modus Ponens
% https://isabelle.in.tum.de/dist/library/FOL/FOL/IFOL.html#IFOL.mp|axiom
is written as
\[
  \ALL P\,Q.\ \ [P\limp Q] \implies ([P] \implies [Q]).
\]
The square brackets (which are omitted in Isabelle theories) represent
an injection from $\tyo$ into $\prop$. % ($[P]$ can be read as “$P$
% holds”)
A consequence of this representation is that every formula of
the object logic appears atomic to $\Meta$.

%% Quantification is handled in the meta-level using a functional $\ALL$
%% with polymorphic type $(\alpha \fun \prop) \fun \prop$. 
Types in Isabelle are organized into \emph{classes} and \emph{sorts};
for ease of exposition, we will omit the former.  The axiomatization
of first-order logic postulates a sort $\{\type{term}\}$ (of
“individuals,” or elements of a first-order universe of discourse) and
stipulates that every further type variable $\alpha$ must be of that
sort. In particular, Isabelle/ZF only postulates one new type $\tyi$
(“sets”) of sort $\{\type{term}\}$. Hence, from the type of the universal
quantifier functional $\forall :: (\alpha \fun \tyo) \fun \tyo$, it
follows that it may only be applied to predicates with a variable of
type $\tyi$. This ensures that the object logic is effectively
first-order.

@@

\subsection{Isabelle/ZF}
\label{sec:isabellezf}

@@

Isabelle/ZF reaches Hessenberg's $|A|\cdot|A| = |A|$. Our decision
(during 2017) to
use this assistant was triggered by its constructibility
library \citep{paulson_2003},
% https://isabelle.in.tum.de/dist/library/ZF/ZF-Constructible/
which contains the development of $L$, the proof that it satisfies
$\AC$, and a version of the Reflection Principle. The latter was
actually encoded as a series of instructions to Isabelle automatic
proof tools that would prove each particular instance of reflection:
This is an example of what was said at the end of Section~\ref{sec:isabelle-metalogic-meta}.

@@

\subsection{Internalized formulas}
\label{sec:internalized-formulas}

\session{ZF-Constructible} defines the set $\formula$ of codes for
first-order formulas. These, alongside with lists, are instances of
Isabelle/ZF treatment of inductively defined (internal) datatypes; induction
and recursion theorems for them are proved automatically (this is in
constrast to general well-founded recursion, for which one has to work
with the fundamental recursor $\isatt{wfrec}$).

To avoid problems related to the binding of variables, de Bruijn
indices \cite{MR0321704} are used instead. The satisfaction predicate
$\isatt{sats}::\tyi\fun\tyi\fun\tyi\fun\tyo$ takes as arguments a set $M$, a list
$\mathit{env}\in\isatt{list}(M)$ for the assignment of free indices,
and $\phi\in\formula$, and it is written
$M,\mathit{env}\models\phi$ in our formalization. This completes the
description of the fourth and last formal layer of the development.

Internalized formulas for most (but not all) of the relational
concepts can be obtained by guiding the automatic tactics. Actually,
great many of the concepts in \session{ZF-Constructible} where
internalized by hand; this is the case for union,
%% \begin{multline*}
%%   \isatt{big{\uscore}union{\uscore}fm}(A,z) \equiv \\
%%   ((\cdot\forall\cdot\cdot0 \in succ(z)\cdot \longleftrightarrow
%%   (\cdot\exists\cdot\cdot0 \in succ(succ(A))\cdot \land \cdot1 \in
%%   0\cdot\cdot\cdot)\cdot\cdot)
%% \end{multline*}
for which we have the following satisfaction lemma:
% https://isabelle.in.tum.de/dist/library/ZF/ZF-Constructible/L_axioms.html#L_axioms.sats_big_union_fm|fact
\begin{multline}\label{eq:sats_big_union_fm}
  x \in \omega \implies y \in \omega \implies \mathit{env} \in \isatt{list}(A)
  \implies \\
  A, \mathit{env} \models \isatt{big{\uscore}union{\uscore}fm}(x,y)%\cdot\union x \isatt{ is } y\cdot
  \longleftrightarrow \\
  \isatt{big{\uscore}union}(\isatt{\#\#} A, \isatt{nth}(x,
  \mathit{env}), \isatt{nth}(y, \mathit{env}))
\end{multline}
Note that $x$ and $y$ above are de Bruijn indices,
$\isatt{nth}(x,\mathit{env})$ is the $x$th element of $\mathit{env}$
and $\isatt{\#\#}A::\tyi\fun \tyo$ is the class corresponding to the
set $A::\tyi$.
